---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: yes
---
```{r global_options, echo=TRUE}
knitr::opts_chunk$set(echo=TRUE,warning=FALSE, message=FALSE)
```
## Loading library and preprocessing the data

```{r}
library(plyr)
library(dplyr)
library(ggplot2)
activity<-read.csv("activity.csv",header = TRUE)
```

```{r}

str(activity)
```
Convert the date variable from Factor to Date variable
```{r}
activity$date<-as.Date(activity$date)

```


## What is mean total number of steps taken per day?
Compute the data needed for the plot
```{r}
tot.steps.day<-group_by(activity,date)%>%summarise(total.steps=sum(steps))
head(tot.steps.day)
```
Plot the data on histogram Q1.1
```{r Histogram Total Number of steps per day}
hist(tot.steps.day$total.steps,
     breaks=50,
     main="Total Number of steps per day",
     xlab="Total steps per day",
     ylab="count",
     col="blue")
```
Calculate the mean and median of total number of steps per day, Q1.2
```{r}
steps.avg.day<-mean(tot.steps.day$total.steps,na.rm = TRUE)
steps.median.day<-median(tot.steps.day$total.steps,na.rm = TRUE)
```
```{r}
steps.avg.day
steps.median.day
```
## What is the average daily activity pattern?
Compute the data needed for the plot
```{r}
avg.steps.interval<-group_by(activity,interval)%>%summarise(average.steps=mean(steps,na.rm=TRUE))
```
Plot the requested data, Q2.1
```{r Time series plot Average Number of steps per interval}
plot(avg.steps.interval$interval,avg.steps.interval$average.steps,
     type="l",
     xlab="Intervals",
     ylab="Average number of steps",
     main="Average Number of Steps per Intervals",
     col="dark green")
```
Interval with max number of steps, Q2.2
```{r}
max.steps.int<-which.max(avg.steps.interval$average.steps)
interval.max.steps<-avg.steps.interval[max.steps.int,1]
```

```{r}
interval.max.steps
```
8:35 is the 5-minute interval on average with the maximum number of steps

## Imputing missing values
Total number of missing values in the dataset
```{r}
table(complete.cases(activity))
```
Create a new data frame and impute missing values using average steps per 5 min-interval 
```{r}
imput.activity<-activity
imput.activity$steps<-ifelse(is.na(imput.activity$steps),
                             avg.steps.interval$average.steps[match(imput.activity$interval,
                                avg.steps.interval$interval)],
                                imput.activity$steps)
```
```{r}
table(complete.cases(imput.activity))
```
Compute total number of steps per day with the imputed missing values
```{r}
imput.tot.steps.day<-group_by(imput.activity,date)%>%summarise(total.steps=sum(steps))
```

```{r}
head(imput.tot.steps.day)
```
Plot the data on histogram, Q3.4
```{r Total number of steps per day (NAs imputed) } 
hist(imput.tot.steps.day$total.steps,
     breaks=50,
     main="Total number of steps per day (NAs imputed)",
     xlab="Total steps per day",
     ylab="count",
     col="red")
```
Report Mean and Median with the imputed missing values
```{r}
imput.steps.avg.day<-mean(imput.tot.steps.day$total.steps)
imput.steps.median.day<-median(imput.tot.steps.day$total.steps)
```

```{r}
imput.steps.avg.day
```

```{r}
imput.steps.median.day
```
The average and median steps by day are not affected by the adopted imputation of the NA values approach.
The distribution of the total nr of steps per day becomes a closer approximation of a Gaussian distribution.

## Are there differences in activity patterns between weekdays and weekends?
Identify the day of the week variable from the imput.activity data frame and then create a weekday or weekend variable
```{r}
imput.activity$day<-weekdays(imput.activity$date)
w.day<-c("Monday","Tuesdya","Wednesday","Thursday","Friday")
w.end<-c("Saturday","Sunday")
imput.activity$weekpart<-ifelse(imput.activity$day %in% w.day,"weekday","weekend")
```

```{r}
head(imput.activity)
```
```{r}
table(imput.activity$weekpart)
```
Group_by data by interval and type of weekday/end 
```{r}
avg.steps.wday.wend<-group_by(imput.activity,interval,weekpart)%>%summarise(average.steps=mean(steps))
```

```{r}
head(avg.steps.wday.wend)
```

Plotting the  weekday and weekend average steps per intervals
```{r}
avg.steps.wday.wend.plot<-ggplot(avg.steps.wday.wend,aes(interval,average.steps))
avg.steps.wday.wend.plot<-avg.steps.wday.wend.plot+geom_line()+facet_grid(weekpart~.)
```
Average activity steps are higher on weekday days than weekend days
```{r Plot average number of steps weekday and weekend}
print(avg.steps.wday.wend.plot)
```